---
title: "FastQC"
author: "Xiting Yan (ported to Bouchet/mouse by jfd42)"
date: "2026-07-06"
output:
  html_notebook:
    code_folding: hide
    fig_caption: yes
    highlight: tango
    number_sections: no
    theme: united
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE,echo = TRUE,cache=TRUE,warning=FALSE,message = FALSE,results='hold',cache.lazy = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap',dev=c('png','postscript'))

library(gdata)
library(knitr)
library(nlme)
library(rgl)
library(gplots)

knit_hooks$set(webgl = hook_webgl)

# --- Bouchet paths (was McCleary /home/xy48/scratch_pi_xy48/xy48/2026_Manning_PAvRVBulk) ---
home.dir<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica_RV"
source("/nfs/roberts/project/pi_epm38/jfd42/pipelines/Xiting_bulkRNA_FastQtoCounts/my_functions.R")
output.dir<-file.path(home.dir,"Results")
```


# Background
In this note, we demonstrate how to use R to generate sh files to run FastQC on the silica PA vs RV bulk RNA-seq data (Plasmidsaurus SS8VZG, mouse). The fastq files are single-end and reside flat in one directory (no per-sample subdirectories). Lung and RV are analyzed separately: point `data.dir` at one tissue folder per run.

# FastQC on raw reads

This will be run on the Bouchet HPC. Each sample gets one sh file; all jobs are submitted via jobsub.bat.

```{r eval=FALSE}
# This chunk needs to be run on the HPC.
# NOTE: set data.dir to the tissue you are processing (lung shown; swap to rv for the RV run).
data.dir<-"/nfs/roberts/project/pi_epm38/jfd42/silica/FASTQ_files/rv"
work.dir<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica_RV"

script.dir<-file.path(work.dir,"scripts_FastQC")
output.dir<-file.path(work.dir,"results_FastQC")

if(file.exists(script.dir)==F){
  dir.create(script.dir)
}

if(file.exists(output.dir)==F){
  dir.create(output.dir)
}

jobsub.filepath<-file.path(script.dir,"jobsub.bat")
if(file.exists(jobsub.filepath)){
  file.remove(jobsub.filepath)
}
file.create(jobsub.filepath)

# All fastq files are flat in data.dir (no per-sample subdirectories)
fastq.filenames<-list.files(data.dir, pattern="\\.fastq\\.gz$", full.names=FALSE)

for(i in 1:length(fastq.filenames)){
  sample.name<-sub("\\.fastq\\.gz$","",fastq.filenames[i])
  script.filepath<-file.path(script.dir,paste(sample.name,".sh",sep=""))
  output.subdir<-file.path(output.dir,sample.name)
  if(file.exists(output.subdir)==F){
    dir.create(output.subdir)
  }

  fastq.filepath<-file.path(data.dir,fastq.filenames[i])

  cmd.out<-"#!/bin/bash\n"
  cmd.out<-paste(cmd.out,"#SBATCH --partition=day\n",sep="")
  cmd.out<-paste(cmd.out,"#SBATCH --job-name=",sample.name,"\n",sep="")
  cmd.out<-paste(cmd.out,"#SBATCH --ntasks=1\n",sep="")
  cmd.out<-paste(cmd.out,"#SBATCH --cpus-per-task=10\n",sep="")
  cmd.out<-paste(cmd.out,"#SBATCH --mem-per-cpu=10G\n",sep="")
  cmd.out<-paste(cmd.out,"#SBATCH --time=1-00:00:00\n",sep="")
  cmd.out<-paste(cmd.out,"module load Perl/5.38.2-GCCcore-13.3.0\n",sep="")
  cmd.out<-paste(cmd.out,"module load FastQC/0.12.1-Java-17\n",sep="")

  temp<-paste("fastqc -o ",output.subdir," -t 10 ",fastq.filepath,"\n",sep="")
  cmd.out<-paste(cmd.out,temp,sep="")
  cat(cmd.out,file=script.filepath,append=F)

  cat("sbatch < ",script.filepath,"\n",sep="",file=jobsub.filepath,append=T)
}
system(paste("chmod 700 ",jobsub.filepath,sep=""))
```

## Generate Summary

Unzip the FastQC result zip files.
```{r}
# Run on HPC
result.dir<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica_RV/results_FastQC"
sample.names<-list.files(result.dir)

for(i in 1:length(sample.names)){
  filenames<-list.files(file.path(result.dir,sample.names[i]))
  filenames<-filenames[grep(".zip",filenames)]
  for(j in 1:length(filenames)){
    unzip(file.path(result.dir,sample.names[i],filenames[j]),exdir=file.path(result.dir,sample.names[i]))
  }
}
```

Extract and merge results across all samples.
```{r}
# Run on HPC
source("/nfs/roberts/project/pi_epm38/jfd42/pipelines/Xiting_bulkRNA_FastQtoCounts/my_functions.R")
result.dir<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica_RV/results_FastQC"
output.filepath<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica_RV/FastQC_results_table.txt"
sample.names<-list.files(result.dir)
cmd.out<-data.frame()
filenames.vect<-character()

for(i in 1:length(sample.names)){
  result.subdir<-file.path(result.dir,sample.names[i])
  temp.dirs<-list.dirs(result.subdir,full.names=F,recursive=FALSE)
  input.filenames<-file.path(temp.dirs,"summary.txt")
  filenames.vect<-c(filenames.vect,temp.dirs)
  input.filepath<-file.path(result.subdir,input.filenames)

  my.data.1<-data.frame()
  for(j in 1:length(input.filepath)){
    temp.text<-as.matrix(read.table(input.filepath[j],sep="\t",header=F,check.names=F))
    temp.values<-temp.text[,1]
    names(temp.values)<-temp.text[,2]
    my.data.1<-rbind(my.data.1,as.data.frame(t(temp.values)))
  }

  input.filenames<-file.path(temp.dirs,"fastqc_data.txt")
  input.filepath<-file.path(result.subdir,input.filenames)
  my.data<-data.frame()
  for(j in 1:length(input.filepath)){
    temp.text<-readLines(input.filepath[j])
    over.start<-grep(">>Overrepresented sequences",temp.text)
    adapter.start<-grep(">>Adapter Content",temp.text)
    my.temp<-data.frame(overrepresented_sequence=my.element.extract(temp.text[over.start+2],splitchar="\t",index=1))
    temp.names<-unlist(strsplit(temp.text[adapter.start+1],split="\t"))
    temp.values<-unlist(strsplit(temp.text[length(temp.text)-1],split="\t"))
    temp.values<-temp.values[-1]
    names(temp.values)<-temp.names[2:length(temp.names)]
    temp.values<-as.data.frame(t(temp.values))
    my.temp<-cbind(my.temp,temp.values)
    my.data<-rbind(my.data,my.temp)
  }

  my.data<-cbind(my.data.1,my.data)
  cmd.out<-rbind(cmd.out,my.data)
}

rownames(cmd.out)<-filenames.vect

cat("filenames\t",file=output.filepath,append=F)
write.table(cmd.out,file=output.filepath,sep="\t",row.names=T,col.names=T,quote=F,append=T)
```
