---
title: "STAR mapping on bulk RNAseq data"
author: "Xiting Yan (ported to Bouchet/mouse by jfd42)"
date: "2026-07-06"
output:
  html_notebook:
    code_folding: hide
    fig_caption: yes
    highlight: tango
    number_sections: no
    theme: united
    toc: yes
    toc_depth: 6
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE,echo = TRUE,cache=TRUE,warning=FALSE,message = FALSE,results='hold',cache.lazy = FALSE)
knitr::opts_knit$set(eval.after = 'fig.cap',dev=c('png','postscript'))

library(gdata)
library(knitr)
library(nlme)
library(rgl)
library(gplots)

knit_hooks$set(webgl = hook_webgl)
```


# Background

In this note, we use R to generate sh files to run STAR on the silica PA vs RV bulk RNA-seq data (Plasmidsaurus SS8VZG). The reads are **single-end** and map to the **mouse genome (GRCm39, GENCODE M39)**.

# Build Genome Index

Build the GRCm39 (GENCODE M39) STAR index if not already available.

**sjdbOverhang note:** this should be (max read length - 1). Xiting's human data maxed at 94 bp, so she used sjdbOverhang=93. CONFIRM your Plasmidsaurus read length before building (from the first FastQC "Sequence length", or:
`zcat one_sample.fastq.gz | head -400 | awk 'NR%4==2{print length($0)}' | sort -n | tail -1`).
If your reads are also 94 bp, keep 93 below. Otherwise change both the index build and the STAR call.

Run the reference download first (00_download_reference.sh) so the FASTA and GTF exist on project space.

```
salloc -t 6:00:00 --mem=120Gb -N 1 -n 1 --cpus-per-task 20 -p day
module load STAR/2.7.11b-GCC-13.3.0
STAR --runMode genomeGenerate --runThreadN 20 \
  --genomeDir /nfs/roberts/project/pi_epm38/jfd42/STARindex/GRCm39_94 \
  --genomeFastaFiles /nfs/roberts/project/pi_epm38/jfd42/GENCODE/GRCm39/GRCm39.primary_assembly.genome.fa \
  --sjdbGTFfile /nfs/roberts/project/pi_epm38/jfd42/GENCODE/GRCm39/gencode.vM39.primary_assembly.annotation.gtf \
  --sjdbOverhang 93
```

# Run STAR mapping

Single-end reads; one fastq file per sample. Set data.dir to the tissue you are processing (lung shown; swap to rv for the RV run). Point data.dir at the trimmed cutadapt output for this tissue.
```{r}
# This chunk needs to be run on the HPC directly
home.dir<-"/nfs/roberts/home/jfd42"
source("/nfs/roberts/project/pi_epm38/jfd42/pipelines/Xiting_bulkRNA_FastQtoCounts/my_functions.R")
work.dir<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica"
data.dir<-file.path(work.dir,"results_cutadapt")   # trimmed reads (one subdir per sample)
star.index.dir<-"/nfs/roberts/project/pi_epm38/jfd42/STARindex/GRCm39_94"
sjdb.filepath<-"/nfs/roberts/project/pi_epm38/jfd42/GENCODE/GRCm39/gencode.vM39.primary_assembly.annotation.gtf"

sample.names<-list.files(data.dir)
sample.names<-sapply(sample.names,my.element.extract,splitchar="\\.",index=1)

script.dir<-file.path(work.dir,"scripts_STAR_GRCm39")
output.dir<-file.path(work.dir,"results_STAR_GRCm39")

if(file.exists(script.dir)==F){
  dir.create(script.dir)
}

if(file.exists(output.dir)==F){
  dir.create(output.dir)
}

jobsub.filepath<-file.path(script.dir,"jobsub.bat")
if(file.exists(jobsub.filepath)){
  file.remove(jobsub.filepath)
}
file.create(jobsub.filepath)

for(i in 1:length(sample.names)){
  data.subdir<-file.path(data.dir,sample.names[i])
  output.subdir<-file.path(output.dir,sample.names[i])
  if(file.exists(output.subdir)==F){
    dir.create(output.subdir)
  }

  # Get the final trimmed fastq (exclude round1/round2 intermediates)
  fastq.filenames<-list.files(data.subdir,full.names=FALSE,recursive=FALSE)
  fastq.filenames<-fastq.filenames[grep("\\.fastq\\.gz$",fastq.filenames)]
  fastq.filenames<-fastq.filenames[-grep("_round1",fastq.filenames)]
  fastq.filenames<-fastq.filenames[-grep("_round2",fastq.filenames)]
  fastq.filepath<-file.path(data.subdir,fastq.filenames)
  if(length(fastq.filepath)>1){
    cat("i=",i,":",sample.names[i]," has more than 1 fastq file.\n")
  }
  jobname<-sample.names[i]

  script.filepath<-file.path(script.dir,paste(jobname,".sh",sep=""))

  temp<-"#!/bin/bash\n"
  temp<-paste0(temp,"#SBATCH --partition=day\n")
  temp<-paste0(temp,"#SBATCH --job-name=",jobname,"\n")
  temp<-paste0(temp,"#SBATCH --nodes=1 --cpus-per-task=10 --mem=40Gb\n")
  temp<-paste0(temp,"#SBATCH --time=1-00:00:00\n")
  temp<-paste0(temp,"#SBATCH --mail-type=NONE\n")
  temp<-paste0(temp,"#SBATCH --error=",script.filepath,".e%J\n","#SBATCH --output=",script.filepath,".o%J","\n",sep="")
  temp<-paste(temp,"module load STAR/2.7.11b-GCC-13.3.0\n",sep="")

  # Single-end: only one file in --readFilesIn
  temp<-paste(temp,"STAR --readFilesCommand zcat --outSAMtype BAM SortedByCoordinate --outSAMprimaryFlag AllBestScore --bamRemoveDuplicatesType UniqueIdentical --outSAMstrandField intronMotif --outFilterIntronMotifs RemoveNoncanonical --runThreadN 10 --genomeDir ",star.index.dir," --sjdbGTFfile ",sjdb.filepath," --sjdbOverhang 93 --twopassMode Basic"," --readFilesIn ",fastq.filepath," --outFileNamePrefix ",file.path(output.subdir,sample.names[i]),sep="")

  cat(temp,file=script.filepath,append=F)
  cat("sbatch < ",script.filepath,"\n",sep="",file=jobsub.filepath,append=T)
}
system(paste("chmod 700 ",jobsub.filepath,sep=""))
```

Submit all jobs by running `jobsub.bat` in the terminal.

# Generate mapping summary table
```{r}
rm(list=ls())
source("/nfs/roberts/project/pi_epm38/jfd42/pipelines/Xiting_bulkRNA_FastQtoCounts/my_functions.R")
work.dir<-"/nfs/roberts/scratch/pi_epm38/jfd42/silica"
data.dir<-file.path(work.dir,"results_STAR_GRCm39")
output.filepath<-file.path(work.dir,"STAR_mapping_summary.txt")

file.names<-list.files(data.dir)
sample.names<-file.names
star.matrix<-numeric()

for(i in 1:length(file.names)){
  filepath<-list.files(file.path(data.dir,file.names[i]))
  filepath<-filepath[grep(".final.",filepath)]
  filepath<-file.path(data.dir,file.names[i],filepath)

  temp<-readLines(filepath)
  temp.vect<-as.numeric(my.element.extract(temp[grep(" Number of input reads",temp)],splitchar="\t",index=-1))
  temp.vect<-c(temp.vect,as.numeric(my.element.extract(temp[grep("Uniquely mapped reads number",temp)],splitchar="\t",index=-1)))
  temp.vect<-c(temp.vect,as.numeric(my.element.extract(temp[grep("Number of reads mapped to multiple loci",temp)],splitchar="\t",index=-1)))
  star.matrix<-rbind(star.matrix,temp.vect)
}

rownames(star.matrix)<-sample.names
colnames(star.matrix)<-c("total","mapped_unique","mapped_multi")
# Single-end: no *2 multiplication (paired-end counts both reads)
star.matrix<-cbind(star.matrix,star.matrix[,2]/star.matrix[,1])
star.matrix<-cbind(rownames(star.matrix),star.matrix)
colnames(star.matrix)<-c("sample_names","total","mapped_unique","mapped_multi","unique_maprate")

write.table(star.matrix,file=output.filepath,append=F,sep="\t",row.names=F,col.names=T,quote=F)
```
